---
title: "lab5"
author: "Alicia Fennell"
date: "10/29/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Lab 5 objectives:

- Learn to parse dates with 'libridate' package
- Get counts of observations with count()
- Uncount() rows 
- One and two sample t-test
- Create a heat map with geom_tile()

```{r, include = FALSE}
#include = FALSE means code and outputs will not show up

library(tidyverse)
library(here)
library(janitor)

# read in data
# lobster data untidy because of "lobster count" ... those are all individual observations but are grouped together as a count. It's better if every observation has its own row and then you can decide how you want to group and/or count things...
# -99999 replace with NA

lobster_abundance <- read_csv(here("data", "lobster_abundance.csv"),
                              na = "-99999") %>% 
  clean_names()
  
```

Use tidyr::uncount() function to convert lobster data from frequency format to case format. (give each observation its own row)... uncount() is great whenever you have a count/frequency column 

```{r}
lobster_tidy <- lobster_abundance %>% 
  uncount(lobster_count)
```

#### Exploratory data visualization 

We'll only consider 'site' as our variable of interest
LOok at data, think about it, try different things, repeat

```{r}

ggplot(lobster_tidy, aes(x = site, y = size_mm)) +
  geom_jitter(aes(color = site),
              alpha = .5,
              width =.2)

# we can start asking questions...
# Let's try histogram, differing by site
# facet_wrap makes them into different plots based on site
# scale = "free" means the scales can differ plot to plot based on the data- dont do this if we are comparing data to each other. but if we want to see them individually, we can scale = free.

ggplot(lobster_tidy, aes(x = size_mm)) +
  geom_histogram(aes(fill = site)) +
  facet_wrap(~site, scale = "free") 

# see how our sample compares to a perfect theorical version...
# either we will see how ALL lobsters compare to perfect population / normally distributed, OR we can add facet_wrap() to see by site how that relationship looks
# geom_qq() defaults to normal distribution. 

ggplot(lobster_tidy, aes(sample = size_mm)) +
  geom_qq() +
  facet_wrap(~site)
```

#### Convert the date column to class 'Date'

"lubridate" package to convert to date format and easily parse month and year. 
We've changed the class to "date."

```{r}
lobster_date <- lobster_tidy %>% 
  mutate(date_new = lubridate::mdy(date))
```

We'll parse year and month using lubridate::month() and year()

```{r}

lobster_parse_date <- lobster_date %>% 
  mutate(obs_month = lubridate::month(date_new, label = TRUE),
         obs_year = lubridate::year(date_new))

# label = TRUE will give month abb instead of number

```

lubridate is COOL!

Now let's find counts of observed lobsters based on different heirarchical groupings

count() function allows you to create counts based on grouping that YOU find most important. Thats why we started with uncount(). We get to decide groupings to count things by.

```{r}
lobster_ym <- lobster_parse_date %>% 
  count(obs_year, obs_month)

lobster_y <- lobster_parse_date %>% 
  count(obs_year)

lobster_site <- lobster_parse_date %>% 
  count(site)

lobster_site
```

If we want to create a summary table that contains stats other than count by group, it's easier to use group_by() + n()

for example, if we are only interested in site...

```{r}

lobster_summary <- lobster_parse_date %>% 
  group_by(site) %>% 
  summarize(lobster_number = n(),
            mean_size = mean(size_mm, na.rm = TRUE),
            sd_size = sd(size_mm, na.rm = TRUE))

lobster_summary
```

#### Find confidence intervals

Use t.test() function to find confidence intervals for one sample and perform t tests to compare means of two samples. 

```{r}
ivee_lobsters <- lobster_tidy %>% 
  filter(site == "IVEE") %>% 
  pull(size_mm)

## pull() makes it into a vector? Not a table ?

t.test(ivee_lobsters)

```

We want to know if there's a significant difference between NAPLES and MOHAWAK... if we think means are normally dist and we think the mean is an appropriate... something... we'll compare those two.

#### 2 sample t test to compare means

We want to ask if there's a sig difference in lobster lengths at these two places. We've done exploratory analyses to determine that 2 sample t test for means comparison is appropriate. 

```{r}

# first make vector for lobster sizes at naples and mohawk

napl_lobsters <- lobster_tidy %>% 
  filter(site == "NAPL") %>% 
  pull(size_mm)

mohk_lobsters <- lobster_tidy %>% 
  filter(site == "MOHK") %>% 
  pull(size_mm)

mn_ttest <- t.test(mohk_lobsters, napl_lobsters)

mn_ttest

```

There's a significant difference in lobster lengths between mohawk and naples reef. Statement of significance should be least interesting part of data. P value isn't that cool. 

```{r}
lobster_mn <- lobster_tidy %>% 
  filter(site %in% c("NAPL", "MOHK"))

mn_ttest2 <- t.test(size_mm ~ site, data = lobster_mn)

mn_ttest2

# Same results! Just different ways to get there
```

If you're writing a report/summary of this data / t test, don't copy and paste values! not reproducible. 

Use ?t.test in the console to see what the names are associated with values

So you can call these things with mn_ttest2$p.value etc... 

Check it out! Here's the p value: `r mn_ttest2$p.value`

When knitted, that will show up as a number. Reproducible.

Here's t statistic: `r mn_ttest2$statistic`

"in line referencing" is a great way to not copy and paste. will automatically make updates for you. 

#### Heat map

generally good for showing 3 variables... 

```{r}

lobster_ys <- lobster_parse_date %>% 
  count(obs_year, site)

ggplot(lobster_ys, aes(x = obs_year,
                       y = site)) +
  geom_tile(aes(fill = n))
```

